---
title: "A paired RNA-seq response in airway cells"
subtitle: "Case 1: Real counts, paired design and DESeq2"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
execute:
  error: false
  warning: true
  message: false
---

## Question and learning outcomes

Does dexamethasone alter gene expression in cultured human airway smooth-muscle cells, accounting for differences between cell lines? This case uses actual processed RNA-seq counts, moving from the transparent hypothetical model in [P09](p09_rnaseq.qmd) to a negative-binomial analysis with DESeq2.

By the end, you should be able to audit a paired design, distinguish library totals from size factors, specify a treatment contrast, interpret effect sizes and false-discovery-rate adjustment, and examine what changes when pairing is omitted. Allow **3–4 hours**, after B02/B06/B12 and P09.

## Study and measurement

The [Himes et al. study](https://doi.org/10.1371/journal.pone.0099625) profiled primary airway smooth-muscle cell lines. The [airway package](https://bioconductor.org/packages/3.20/data/experiment/html/airway.html) provides the eight untreated/dexamethasone libraries from four cell lines. Each line contributes one sample in each condition. The experiment concerns cultured cells; these eight libraries are not eight independently sampled patients.

The local extract contains the complete package count matrix and sample identifiers, from **airway 1.26.0**. No biological download is needed to run this lesson. Counts depend on the package's gene models and read-counting procedure. Starting from a processed matrix lets us examine statistical analysis, but does not recheck FASTQ quality, mapping or strandedness. Provenance and reuse terms are in [data notes](../case_data.qmd).

The working model is:

$$K_{gj}\sim\mathrm{NB}(\mu_{gj},\alpha_g),\qquad
\log\mu_{gj}=\log s_j+\beta_{g0}+\beta_{g,\mathrm{cell}(j)}+
\beta_{g,\mathrm{dex}}I(j\text{ treated}).$$

Here $s_j$ is a sample size factor and $\alpha_g$ is gene-specific dispersion. The treatment coefficient represents the within-line comparison under the additive design. There is only one library per line–condition combination, so a saturated line-by-treatment interaction is not estimable with replicated residual error.

## Set up once

This real case requires **DESeq2**, unlike the base-R P01–P31 practicals. Install it using the Bioconductor release appropriate to your R version:

```r
install.packages("BiocManager")
BiocManager::install("DESeq2", ask = FALSE, update = FALSE)
```

The worked analysis was checked with R 4.4.1 and DESeq2 1.46.0 (Bioconductor 3.20). Later compatible versions may give slightly different numerical results. Installation is a separate student setup step; rendering does not install packages or retrieve data.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
source(file.path(helper_dir, "case_helpers.R"))
```


```{r}
if (!suppressWarnings(requireNamespace("DESeq2", quietly = TRUE)))
  stop("Install DESeq2 with BiocManager before running Case 1.")
suppressWarnings(suppressPackageStartupMessages(library(DESeq2)))
counts <- case_matrix(case_read("airway_counts.csv.gz"))
meta <- case_read("airway_samples.csv")
stopifnot(!anyDuplicated(meta$sample_id), setequal(colnames(counts), meta$sample_id))
meta <- meta[match(colnames(counts), meta$sample_id), ]
rownames(meta) <- meta$sample_id
meta$cell <- factor(meta$cell)
meta$dex <- factor(meta$dex, levels = c("untrt", "trt"))
stopifnot(identical(colnames(counts), rownames(meta)),
          all(table(meta$cell, meta$dex) == 1),
          all(counts >= 0), all(counts == round(counts)))
knitr::kable(meta, row.names = FALSE)
dim(counts)
```

**Before fitting:** identify the four pairs. Explain why arranging a table in a visually convenient order is not enough to guarantee that metadata match count columns. The identity check must compare explicit sample identifiers.

## Filter, fit and define the contrast

We retain genes with at least 10 counts in at least four libraries. This is a stated teaching prefilter, applied without examining treatment effects. It removes extremely sparse features to improve computation and interpretability; it is not a universal threshold for RNA-seq studies.

```{r}
keep <- rowSums(counts >= 10) >= 4
dds <- DESeqDataSetFromMatrix(countData = counts[keep, ],
                             colData = meta, design = ~ cell + dex)
stopifnot(qr(model.matrix(~ cell + dex, meta))$rank == 5)
dds <- DESeq(dds, quiet = TRUE)
res <- results(dds, contrast = c("dex", "trt", "untrt"), alpha = 0.05)
data.frame(input_genes = nrow(counts), retained_genes = sum(keep),
           tested_pvalues = sum(!is.na(res$pvalue)),
           adjusted_pvalues = sum(!is.na(res$padj)),
           padj_below_005 = sum(res$padj < 0.05, na.rm = TRUE))
```

The contrast is **treated minus untreated** on the log2 scale. A positive log2 fold change corresponds to higher expected normalized abundance in treated cells under this model. DESeq2 estimates dispersions across genes and applies multiple-testing adjustment, including its default independent filtering at the supplied `alpha`. Some adjusted p-values can therefore be missing even when a raw p-value exists. An outlier rule can also affect a gene's test. Inspect these cases rather than changing missing values to 1 or 0.

## Inspect normalization and sample structure

```{r}
#| fig-cap: "Variance-stabilized airway counts: blue untreated, orange dexamethasone; lines connect each cell-line pair. PCA describes variation and does not test treatment."
#| fig-alt: "Variance-stabilized airway counts: blue untreated, orange dexamethasone; lines connect each cell-line pair. PCA describes variation and does not test treatment."
data.frame(sample = meta$sample_id, cell = meta$cell, dex = meta$dex,
           raw_total = colSums(counts), size_factor = sizeFactors(dds))
v <- vst(dds, blind = FALSE)
pc <- prcomp(t(assay(v)), center = TRUE, scale. = FALSE)
explained <- 100 * pc$sdev^2 / sum(pc$sdev^2)
cols <- c(untrt = "#2878a1", trt = "#c55a11")
plot(pc$x[, 1:2], col = cols[as.character(meta$dex)], pch = 19,
     xlab = sprintf("PC1 (%.1f%%)", explained[1]),
     ylab = sprintf("PC2 (%.1f%%)", explained[2]))
for (lev in levels(meta$cell)) {
  ix <- which(meta$cell == lev)
  lines(pc$x[ix, 1], pc$x[ix, 2], col = "grey65")
}
legend("top", legend = c("Untreated", "Dexamethasone"),
       col = cols, pch = 19, bty = "n")
```

Lines connect the samples from each cell line. PCA summarizes variation after a variance-stabilizing transformation; it does not test the treatment contrast. `blind = FALSE` uses the fitted design to estimate the transformation and is not equivalent to removing cell-line or treatment effects. The PCA still contains both. Size factors need not be proportional to total reads because composition matters; see the [composition figure](../visual_guide.qmd#why-is-library-size-not-the-whole-normalization-problem).

## Read effects and uncertainty together

```{r}
#| fig-cap: "Airway MA plot: average normalized abundance versus the unshrunken treated-minus-untreated log2 effect. Highlighting uses adjusted p-values below 0.05."
#| fig-alt: "Airway MA plot: average normalized abundance versus the unshrunken treated-minus-untreated log2 effect. Highlighting uses adjusted p-values below 0.05."
tab <- as.data.frame(res)
tab$gene_id <- rownames(tab)
tab <- tab[order(tab$padj, na.last = TRUE), ]
knitr::kable(head(tab[, c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "padj")], 10),
             digits = 3, row.names = FALSE)
plotMA(res, alpha = 0.05, ylim = c(-5, 5))
```

The MA plot relates average normalized abundance to the estimated treatment effect. The displayed fold changes are **unshrunken model estimates**; for unstable low-count genes, a compatible shrinkage method can be an advanced extension. A large effect with a large standard error is different evidence from a precise moderate effect. FDR adjustment concerns a testing procedure across genes, not the probability that an individual biological mechanism is correct.

Choose a gene **before inspecting its result** for your report, using a documented identifier and annotation version. If you instead choose the first row above, label it an exploratory selected gene. Plot its normalized counts by condition with the four pairs connected. Neither strategy establishes treatment efficacy in patients.

## Sensitivity: omit the cell-line term

This is a comparison with an analysis that discards known design information, not a replacement primary model. Keep the same genes and normalization approach, then assess effects and uncertainty separately.

```{r}
#| fig-cap: "The same retained genes under paired and unpaired models. The diagonal marks equal log2 effects; inspect uncertainty separately."
#| fig-alt: "The same retained genes under paired and unpaired models. The diagonal marks equal log2 effects; inspect uncertainty separately."
unpaired <- DESeqDataSetFromMatrix(countData = counts[keep, ], colData = meta, design = ~ dex)
unpaired <- DESeq(unpaired, quiet = TRUE)
ru <- results(unpaired, contrast = c("dex", "trt", "untrt"), alpha = 0.05)
comparison <- data.frame(gene = rownames(res), paired_effect = res$log2FoldChange,
                        unpaired_effect = ru$log2FoldChange,
                        paired_se = res$lfcSE, unpaired_se = ru$lfcSE)
summary(comparison$unpaired_se / comparison$paired_se)
plot(comparison$paired_effect, comparison$unpaired_effect, pch = 16, cex = 0.3,
     xlab = "Paired log2 fold change", ylab = "Unpaired log2 fold change")
abline(0, 1, col = "#c55a11", lwd = 2)
```

Omitting pairing changes the fitted model and the treatment uncertainty. In a balanced design, effect estimates may agree more closely than standard errors. A larger discovery count alone does not validate a model: the primary design was chosen from how samples were collected, before seeing p-values.

## Exercises and suggested reasoning

1. Describe the experimental unit and why the treatment contrast is paired. What additional replication would support a line-by-treatment interaction?
2. Explain one reason that a gene has a raw p-value but no adjusted p-value. Locate such a gene and check the relevant filtering/outlier information.
3. Change the count prefilter to 10 counts in at least three libraries. Record the number retained and explain why the resulting tested universe differs. Do not choose the threshold by maximizing discoveries.
4. Compare a prespecified gene with an exploratory top gene. Write one bounded conclusion for each.

::: {.callout-tip collapse="true" title="Suggested reasoning"}
Each line supplies its own baseline, so the contrast compares conditions within lines. More independently prepared observations within line–condition combinations would support an interaction assessment, though generalization across donors still requires adequate donors. Independent filtering may omit low-information genes from adjustment; outlier-related missing tests require separate diagnosis. A less strict prefilter changes the eligible features and may alter adjustment as well as computation. Prespecification and exploratory selection have different interpretation, even if the same gene is eventually discussed.
:::

## Report and reproduce

Submit the design, prefilter, contrast, normalization diagnostic, effect/uncertainty table and one sensitivity analysis using the [Case 1 assignment](../assessment.qmd#case-1-a-paired-transcriptomic-response). Include the [input manifest](../data/cases/manifest.csv), annotation limitations and package versions.

```{r}
case_session("DESeq2")
```

For method details, consult the [DESeq2 vignette](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html). Continue to [Case 2](c02_proteomics.qmd) to compare count evidence with protein-enrichment intensities.
