---
title: "Biological databases and sequence retrieval"
subtitle: "Practical 1: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Identify record levels and provenance; retrieve local teaching records by explicit accession; audit one-to-many annotation.

## Biological question and data

Which biological records are needed to connect a measured feature to its sequence and annotation? This offline exercise uses fictional records so retrieval is reproducible and does not depend on a live service.

Records are distinct objects; study, sample, gene and transcript identifiers are not interchangeable. Real retrieval needs species, assembly, version and access checks.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Inspect the resource catalogue

Use the record type to determine what a query returns.

```{r}
catalogue <- data.frame(accession=c("study_demo","sample_demo1","gene_demo1","tx_demo1"),
 type=c("study","sample","gene","transcript"),
 parent=c(NA,"study_demo",NA,"gene_demo1"),version=c("v1","v1","v2","v1"))
catalogue
stopifnot(!anyDuplicated(catalogue$accession))
```

All identifiers are fictional. A sample belongs to a study; a transcript relates to a gene. Neither relationship supplies the measurement values.

## 2. Retrieve and preserve records

Select explicitly and round-trip a temporary CSV. This represents retrieval from a supplied snapshot, not an API call.

```{r}
selected <- catalogue[match(c("gene_demo1","tx_demo1"),catalogue$accession),]
stopifnot(!anyNA(selected$accession))
file <- tempfile(fileext=".csv")
write.csv(selected,file,row.names=FALSE,na="")
restored <- read.csv(file,na.strings="",stringsAsFactors=FALSE)
stopifnot(identical(as.character(restored$accession),selected$accession))
data.frame(accession=restored$accession,version=restored$version,
 snapshot_checksum=unname(tools::md5sum(file)))
unlink(file)
```

A checksum captures this snapshot’s bytes. The version remains separate because a checksum is not a biological accession.

## 3. Audit annotation expansion

A gene-to-transcript mapping can expand the rows. Count distinct biological keys after the join.

```{r}
measurement <- data.frame(gene=c("gene_demo1","gene_demo2"),value=c(8,10))
map <- data.frame(gene=c("gene_demo1","gene_demo1"),transcript=c("tx1","tx2"))
joined <- merge(measurement,map,all.x=TRUE)
joined
stopifnot(nrow(joined)==3,length(unique(joined$gene))==2)
```

Three joined rows still represent two measured genes, not three independent measurements.

## Interpretation and limits

For a real GEO analysis, distinguish series, samples and platforms using [NCBI’s record documentation](https://www.ncbi.nlm.nih.gov/geo/info/datasets.html). This practical does not download or validate an organism-specific annotation.

## Exercises

1. Add an unmatched query and require an informative check.
2. Explain why transcript rows cannot be counted as independent gene measurements.
3. What extra fields would a variant-annotation snapshot need?

::: {.callout-note collapse="true" title="Suggested answers"}
1. `match` returns NA; stop or record the exclusion before indexing.
2. Both transcript annotations reuse the gene measurement.
3. Species, assembly, coordinate convention, counted alleles where relevant, annotation release, source and retrieval details.
:::

## Continue learning

[B03 Databases](../notes/b03_databases.qmd) and [B04 Programming](../notes/b04_programming.qmd).

Continue with [sequence handling](p02_sequences.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
