---
title: "Basic sequence analysis and manipulation"
subtitle: "Practical 2: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Parse FASTA records; validate sequence symbols; calculate called-base GC content; construct a reverse complement.

## Biological question and data

How do file structure and unknown bases affect a sequence summary?

FASTA inputs contain only A, C, G, T and N; these teaching records have no quality scores. N is unknown and is excluded from the GC denominator.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Parse explicit records

Use the local teaching parser, which rejects missing headers, duplicate IDs and invalid bases.

```{r}
source("../R/sequence_helpers.R")
sequences <- read_teaching_fasta(c(">demo1","ACGTNN","GCGC",">demo2","TTTACG"))
sequences
stopifnot(nchar(sequences["demo1"])==10)
```

Wrapped sequence lines are joined within their records. Sequence length includes unknown bases.

## 2. Calculate called-base summaries

Count unknowns separately. An all-unknown sequence has no estimable called-base GC proportion.

```{r}
summary_one <- function(s) {
 x <- strsplit(s,"")[[1]]; called <- x%in%c("A","C","G","T")
 c(length=length(x),called=sum(called),unknown=sum(!called),
 GC=if(any(called))mean(x[called]%in%c("G","C")) else NA_real_)
}
result <- t(vapply(sequences,summary_one,numeric(4)))
result
stopifnot(abs(result["demo1","GC"]-6/8)<1e-12,
 is.na(summary_one("NNN")["GC"]))
```

demo1 has ten characters but eight called bases; six called bases are G or C.

## 3. Reverse-complement the sequence

Reverse the base order as well as complementing each symbol. Keep unknown N as N.

```{r}
reverse_complement <- function(s) {
 x <- rev(strsplit(s,"")[[1]])
 paste(unname(c(A="T",C="G",G="C",T="A",N="N")[x]),collapse="")
}
rc <- vapply(sequences,reverse_complement,character(1))
rc
stopifnot(identical(unname(vapply(rc,reverse_complement,character(1))),unname(sequences)))
```

Reverse complementing twice restores the original sequence, a useful orientation check.

## Interpretation and limits

Called-base GC is a descriptive proportion, not sequencing quality or coding potential. Unknown bases and selected regions can make whole-record comparisons misleading. This parser is intentionally small, not a general FASTA library.

## Exercises

1. Recalculate GC including N in the denominator; explain the change.
2. Add a duplicate header.
3. Why does complementing without reversing give a different object?

::: {.callout-note collapse="true" title="Suggested answers"}
1. It decreases to 6/10, treating unknowns as if they were called non-GC bases.
2. The parser rejects the ambiguous identifier.
3. A reverse complement changes strand orientation; complement alone preserves the original character order.
:::

## Continue learning

[B04 Programming](../notes/b04_programming.qmd) and [B05 Genomics](../notes/b05_genomics.qmd).

Continue with [sequence similarity](p04_similarity.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
