---
title: "Command-line bioinformatics and reproducible execution"
subtitle: "Practical 3: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Pass an explicit input path to a script; inspect exit status; distinguish stdout and generated files; quote paths containing spaces.

## Biological question and data

Can the same file-processing task run independently of an interactive R session?

This portable exercise launches the Rscript executable belonging to the current R installation. It requires no Unix shell or external bioinformatics program. It illustrates the command-line interface pattern used by larger tools.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Create a small input and worker script

Both files live in a fresh temporary directory. The script takes exactly one input argument and validates it.

```{r}
work <- tempfile("cli-example-"); dir.create(work)
input <- file.path(work,"sample counts.csv")
write.csv(data.frame(sample=c("s1","s2","s3"),count=c(10,20,30)),input,row.names=FALSE)
worker <- file.path(work,"summarise.R")
writeLines(c('args <- commandArgs(trailingOnly=TRUE)',
 'stopifnot(length(args)==1L, file.exists(args[1]))',
 'd <- read.csv(args[1])',
 'stopifnot(all(c("sample","count") %in% names(d)))',
 'cat(sum(d$count), "\n")'),worker)
readLines(worker)
```

Explicit arguments make the worker independent of objects in the current global environment.

## 2. Run and capture the process

Quote each path as a separate argument, including the intentionally spaced input filename.

```{r}
executable <- file.path(R.home("bin"),if(.Platform$OS.type=="windows") "Rscript.exe" else "Rscript")
output <- system2(executable,args=c("--vanilla",shQuote(worker),shQuote(input)),
 stdout=TRUE,stderr=TRUE)
status <- attr(output,"status"); if(is.null(status)) status <- 0L
c(status=status,total=as.numeric(tail(output,1)))
stopifnot(status==0L,as.numeric(tail(output,1))==60)
```

The worker reports 60 and exits successfully. Locale-startup messages may appear before the final number on some systems.

## 3. Describe the workflow boundary

Record executable, arguments, input and expected output before applying this pattern to another tool.

```{r}
record <- data.frame(step="sum counts",executable=basename(executable),
 input=basename(input),output="stdout total",expected_total=60)
record
```

For a real aligner, include reference, version, options, resource requirements and the meaning of each output. A zero exit code does not certify biological correctness.

## Interpretation and limits

Use tool-specific documentation for real aligners or quantifiers; this example runs no alignment. Keep restricted data out of command logs and public repositories. Temporary files are left in R’s temporary directory for inspection.

## Exercises

1. What changes if the worker relies on an interactive variable?
2. Why quote the filename?
3. Which check remains necessary after a successful exit?

::: {.callout-note collapse="true" title="Suggested answers"}
1. A clean Rscript process does not contain that variable.
2. Spaces must stay within one argument.
3. Check output structure, sample/reference identity and scientific plausibility, not just process completion.
:::

## Continue learning

[B04 Programming](../notes/b04_programming.qmd) and [B17 Integrity](../notes/b17_integrity.qmd).

Continue with [data handling](p14_data_handling.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
