---
title: "Sequence similarity and evidence for homology"
subtitle: "Practical 4: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Compute aligned identity with a stated denominator; explore a composition-preserving shuffle comparison; distinguish similarity, homology and function.

## Biological question and data

Does a high number of matching bases provide enough evidence to claim common ancestry or shared function?

Two supplied short sequences are compared at corresponding positions without gaps. A shuffle calculation conditions on one sequence’s base composition; it is a restricted teaching null, not BLAST or an evolutionary test.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Measure identity

The sequences have equal length. Record both matches and denominator.

```{r}
a <- strsplit("ACGTACGTACGTACGT","")[[1]]
b <- strsplit("ACGTACGTTCGTACGA","")[[1]]
stopifnot(length(a)==length(b))
observed <- sum(a==b)
c(matches=observed,length=length(a),identity=observed/length(a))
stopifnot(observed==14)
```

Fourteen of sixteen corresponding bases match in this supplied alignment.

## 2. Compare a restricted shuffle null

Shuffle the second sequence without changing composition. Include the usual plus-one Monte Carlo tail convention.

```{r}
set.seed(401)
shuffled <- replicate(499,sum(a==sample(b)))
p_mc <- (1+sum(shuffled>=observed))/(length(shuffled)+1)
c(shuffle_mean=mean(shuffled),observed=observed,shuffle_tail=p_mc)
hist(shuffled,breaks=seq(-.5,16.5,1),xlab="Matches under composition shuffle",main="Restricted teaching null")
abline(v=observed,col="#d95f02",lwd=2)
```

This comparison shows what positional matches look like after a particular randomization. It ignores database size, local-alignment selection, substitution models and sequence dependence.

## 3. Separate claims

Place each claim beside the extra evidence it needs.

```{r}
claims <- data.frame(claim=c("sequence similarity","homology","shared function"),
 evidence=c("explicit aligned comparison","ancestry inference with appropriate context",
 "functional/domain/experimental evidence"))
claims
```

Homology is a relationship of common ancestry, not 87.5% homology. Shared function is a further biological claim.

## Interpretation and limits

Use the [NCBI BLAST glossary](https://www.ncbi.nlm.nih.gov/books/NBK62051/) to distinguish identity and homology. The toy shuffle tail is not a database-search E-value and cannot be substituted for one.

## Exercises

1. Compare two all-A sequences. Why is high identity weak evidence here?
2. What is ignored by positional comparison when lengths differ?
3. Does a small shuffle tail establish function?

::: {.callout-note collapse="true" title="Suggested answers"}
1. Low complexity makes agreement unsurprising; the composition-preserving shuffle gives the same matches.
2. Alignment, indels and local-versus-global coverage must be specified.
3. No; it addresses only the stated restricted null.
:::

## Continue learning

[B05 Genomics](../notes/b05_genomics.qmd).

Continue with [alignment](p05_alignment.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
