---
title: "Protein structure and structural bioinformatics"
subtitle: "Practical 7: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Interpret coordinate units and residue identifiers; calculate pairwise distances and non-neighbour contacts; distinguish geometric contacts from functional evidence.

## Biological question and data

Which residue pairs lie close together in a supplied protein-coordinate model?

The six coordinates are invented C-alpha positions in angstroms, not a protein database structure. Euclidean proximity is calculated under a supplied cutoff; consecutive residues are excluded from the contact summary.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Inspect coordinates and distances

Rows represent residues. Distance is calculated in the stated coordinate system.

```{r}
xyz <- rbind(c(0,0,0),c(3.8,0,0),c(7.6,0,0),c(7.6,3.8,0),c(3.8,3.8,0),c(0,3.8,0))
rownames(xyz) <- paste0("residue",1:6)
D <- as.matrix(dist(xyz))
round(D,2)
stopifnot(abs(D[1,2]-3.8)<1e-12,isSymmetric(D))
```

The folded-back hypothetical geometry places some non-neighbour positions near each other.

## 2. Define a contact rule

A distance below five angstroms is a teaching cutoff. Exclude immediate sequence neighbours and diagonal entries.

```{r}
contact <- D<5 & abs(row(D)-col(D))>1
pairs <- which(contact & upper.tri(contact),arr.ind=TRUE)
data.frame(residue1=rownames(D)[pairs[,1]],residue2=colnames(D)[pairs[,2]],
 distance_A=D[pairs])
stopifnot(nrow(pairs)==2)
image(1:6,1:6,contact,xlab="Residue index",ylab="Residue index",
 col=c("white","#1b9e77"),main="Supplied geometric contact rule")
```

Contacts describe geometry under the chosen threshold. They are not measured interactions or binding energies.

## 3. Assess threshold dependence

Count contacts at several cutoffs using the same non-neighbour rule.

```{r}
cutoffs <- c(4,5,8)
data.frame(cutoff_A=cutoffs,contacts=vapply(cutoffs,function(cut)
 sum((D<cut & abs(row(D)-col(D))>1)[upper.tri(D)]),integer(1)))
```

A contact list changes when its operational definition changes.

## Interpretation and limits

For an actual structure, record accession, model, chain, residue numbering, units, experimental/predicted status and structural uncertainty. A C-alpha cutoff is an abstraction; side chains, ligands, solvent and conformational dynamics affect biological interpretation.

## Exercises

1. What changes when coordinates are converted to nanometres?
2. Why exclude sequential neighbours?
3. Does a contact establish functional necessity?

::: {.callout-note collapse="true" title="Suggested answers"}
1. Values divide by ten and the cutoff must also divide by ten.
2. They reflect backbone proximity and can obscure non-local geometry.
3. No; perturbation and functional evidence are needed.
:::

## Continue learning

[B08 Proteomics and structure](../notes/b08_proteomics.qmd).

Continue with protein abundance (in development). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
