---
title: "Sequencing data quality assessment"
subtitle: "Practical 8: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Decode Phred+33 qualities; calculate per-read expected errors; distinguish read, library and biological replication; apply an explicit QC rule.

## Biological question and data

How do low-quality positions influence a small read-quality summary?

The fictional FASTQ records use ASCII Phred+33 encoding. The error probabilities are interpreted as calibrated per-base probabilities solely for this example. The practical does not assess mapping or a real sequencing run.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Decode quality strings

Confirm the sequence and quality strings have matching lengths.

```{r}
reads <- c(read1="ACGTACGT",read2="ACGTACGT",read3="ACGTNNGT")
quality <- c(read1="IIIIIIII",read2="IIII!!II",read3="55555555")
stopifnot(identical(names(reads),names(quality)),all(nchar(reads)==nchar(quality)))
Q <- t(vapply(quality,function(q) utf8ToInt(q)-33,numeric(8)))
Q
stopifnot(Q[1,1]==40,Q[2,5]==0,Q[3,1]==20)
```

I encodes Q40, ! Q0 and 5 Q20 under the stated convention. The convention must not be guessed from appearance.

## 2. Summarise errors and unknown calls

Sum encoded error probabilities and count unknown bases separately.

```{r}
expected_errors <- rowSums(10^(-Q/10))
unknown <- vapply(reads,function(s) sum(strsplit(s,"")[[1]]=="N"),integer(1))
qc <- data.frame(read=names(reads),mean_Q=rowMeans(Q),expected_errors,unknown)
qc
stopifnot(abs(expected_errors[1]-.0008)<1e-12,expected_errors[2]>2)
```

Two Q0 positions dominate the second read’s expected-error total. Mean quality alone compresses their location and severity.

## 3. Apply and report a specified rule

The supplied rule retains reads with expected errors below .1 and no unknown calls; it is illustrative, not a universal threshold.

```{r}
qc$keep <- qc$expected_errors<.1 & qc$unknown==0
qc
c(input_reads=nrow(qc),retained=sum(qc$keep))
stopifnot(sum(qc$keep)==1)
```

The third read meets the encoded-error rule but fails the unknown-base rule. Report the exclusions and their reason.

## Interpretation and limits

Quality encoding, adapter content, length distributions, contaminants, mapping, coverage and sample identity matter in a complete sequencing workflow. A threshold chosen after looking at a desired result can bias downstream analysis. More retained reads do not create more biological replicates.

## Exercises

1. Why can mean Q hide a problematic read?
2. Does summing error probabilities require independent errors?
3. What information would mapping QC add?

::: {.callout-note collapse="true" title="Suggested answers"}
1. A few extremely low-quality positions can be diluted in the average.
2. No; expectation is additive. Independence matters for some probability calculations, not this sum.
3. Reference alignment, multimapping, coverage and mapping-rate patterns, each requiring interpretation.
:::

## Continue learning

[B05 Genomics](../notes/b05_genomics.qmd) and [B06 Transcriptomics](../notes/b06_transcriptomics.qmd).

Continue with RNA-seq (in development). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
