---
title: "RNA-seq counts, contrasts and uncertainty"
subtitle: "Practical 9: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Align counts and metadata; distinguish known exposure from estimated normalization; fit a restricted count model and interpret log2 effects, dispersion and multiplicity.

## Biological question and data

Does expression differ between the two diets in a hypothetical 48-animal experiment?

The fixture simulates negative-binomial counts for 24 genes, with independent animals, crossed diet/batch and known multiplicative exposure. The first four genes have a supplied diet effect. Gene01 also has a genotype effect. A quasi-Poisson fit below is an intentionally restricted mean-model illustration; its single variance multiplier per gene is not the generating negative-binomial variance function.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Load and align the count panel

Count rows are genes and columns are animals. Metadata order is verified before modelling.

```{r}
source("../R/teaching_data.R")
d <- teaching_data()
meta <- d$metadata; counts <- d$counts
stopifnot(identical(colnames(counts),meta$animal),!anyDuplicated(meta$animal))
meta$exposure <- unname(d$exposure[meta$animal])
stopifnot(all(meta$exposure>0))
counts[1:6,1:8]
xtabs(~diet+batch,meta)
```

Crossed treatment and batch supply separate comparisons. These 24-feature totals are an illustrative panel, not a whole-transcriptome library.

## 2. Inspect exposure-adjusted values

Divide by the supplied exposure, not by a normalization factor estimated from this panel. Plot on a descriptive log2 scale.

```{r}
normalized <- sweep(counts,2,meta$exposure,"/")
stopifnot(max(abs(sweep(normalized,2,meta$exposure,"*")-counts))<1e-10)
boxplot(log2(normalized[1,]+1)~meta$diet,col=c("#86B7A3","#E5A66B"),
 xlab="Diet",ylab="log2(exposure-adjusted gene01 count + 1)")
c(raw_panel_total_range=range(colSums(counts)))
```

The +1 transformation supports a plot; it is not the response used in the count model. Real RNA-seq normalization must consider composition and cannot assume this known exposure.

## 3. Fit the mean model and report uncertainty

Fit each gene with diet, batch and a log-exposure offset. The diet coefficient uses the natural-log link; divide it and its standard error by log(2) to obtain log2 units.

```{r}
fit_gene <- function(y) {
 fit <- glm(y~diet+batch+offset(log(exposure)),data=meta,family=quasipoisson())
 tab <- coef(summary(fit))
 c(log2_effect=tab["dietsupplement","Estimate"]/log(2),
 SE_log2=tab["dietsupplement","Std. Error"]/log(2),
 p_value=tab["dietsupplement","Pr(>|t|)"],
 variance_multiplier=summary(fit)$dispersion)
}
result <- as.data.frame(t(apply(counts,1,fit_gene)))
result$BH <- p.adjust(result$p_value,"BH")
result$lower <- result$log2_effect-qt(.975,nrow(meta)-3)*result$SE_log2
result$upper <- result$log2_effect+qt(.975,nrow(meta)-3)*result$SE_log2
result[1:8,]
stopifnot(all(is.finite(as.matrix(result))),all(result$BH>=result$p_value-1e-12))
```

The intervals and p-values are conditional on this restricted working model. Genotype variation for gene01 remains unmodelled here. A fitted association and an approximate p-value are not a causal-mechanism claim.

## Interpretation and limits

This practical teaches design, exposure, contrasts and uncertainty; it is not a production differential-expression pipeline. The [DESeq2 workflow](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) adds composition-aware normalization and a gene-level negative-binomial dispersion framework. Gene filtering, relatedness, batch design and larger-scale validation need explicit decisions for real data.

## Exercises

1. What supplied log2 effect corresponds to .65 on the natural-log scale?
2. Why is Poisson variance too restrictive for these generated counts?
3. Change the seed and explain why estimates and adjusted p-values change.
4. What would a complete treatment–batch confounding problem prevent?

::: {.callout-note collapse="true" title="Suggested answers"}
1. `.65/log(2)`, about .94; a finite-sample estimate need not equal it.
2. The generating variance is mu + mu squared/12, greater than mu.
3. Sampling realization changes, while the generating parameters remain the same.
4. Separate estimation of diet and batch effects.
:::

## Continue learning

[B02 Design](../notes/b02_design.qmd), [B06 Transcriptomics](../notes/b06_transcriptomics.qmd) and [B12 Statistics](../notes/b12_statistics.qmd).

Continue with [pathway interpretation](p10_pathways.qmd), or use the Multi-Omics extension on [assay QC](p16_multiomics_qc.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
