---
title: "Functional annotation and pathway enrichment"
subtitle: "Practical 10: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Preserve feature-to-gene mappings; define an eligible background; calculate a hypergeometric tail and verify its equivalent Fisher test; assess background sensitivity.

## Biological question and data

Is a selected gene list enriched for a supplied pathway under an explicitly defined selection null?

The gene/pathway table is fictional. Every background gene is eligible, and the null treats a fixed-size selected set as uniformly drawn from that background. Selected genes are supplied for teaching, not discovered by a calibrated genome-wide analysis.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Define the eligible universe

Use the 24 assayed gene IDs and a prespecified six-gene list.

```{r}
source("../R/teaching_data.R")
d <- teaching_data()
background <- rownames(d$counts)
selected <- sprintf("gene%02d",1:6)
stopifnot(!anyDuplicated(background),all(selected%in%background))
annotation <- d$annotation
stopifnot(!anyDuplicated(annotation$gene),setequal(annotation$gene,background))
table(annotation$pathway)
```

The measured universe is known. Counting annotations outside it would change the null population.

## 2. Calculate enrichment

For each set, compute overlap, null expectation and an upper-tail hypergeometric p-value. Confirm the first result with a 2-by-2 Fisher test.

```{r}
one_set <- function(pathway) {
 member <- annotation$gene[annotation$pathway==pathway]
 N <- length(background); K <- length(member); n <- length(selected)
 x <- sum(selected%in%member)
 c(N=N,K=K,n=n,x=x,expected=n*K/N,p=phyper(x-1,K,N-K,n,lower.tail=FALSE))
}
result <- as.data.frame(t(vapply(unique(annotation$pathway),one_set,numeric(6))))
result$BH <- p.adjust(result$p,"BH")
result
z <- result[1,]
tab <- matrix(c(z$x,z$n-z$x,z$K-z$x,z$N-z$K-z$n+z$x),2,byrow=TRUE)
stopifnot(abs(fisher.test(tab,alternative="greater")$p.value-z$p)<1e-12)
```

PathwayA has six selected members in a background of eight pathway members. The enrichment model concerns gene-set representation, not measured pathway activity.

## 3. Change the background and explain the effect

Keep the pathway count fixed while adding unmeasured nonmembers to the universe. This is deliberately an inappropriate comparison for the same assay.

```{r}
data.frame(background=c("eligible panel","inflated universe"),
 N=c(24,100),p=c(phyper(5,8,16,6,lower.tail=FALSE),
 phyper(5,8,92,6,lower.tail=FALSE)))
```

Adding many nonmembers makes the same overlap appear more surprising. That is a change in assumptions, not more biological evidence.

## Interpretation and limits

Hypergeometric enrichment assumes an appropriate equal-selection null. Expression level, feature length, detectability, correlation and annotation overlap can require a more suitable method. Document mapping, set version, selection rule, background and multiplicity.

## Exercises

1. Why exclude unmeasured genes from this universe?
2. Does a small p-value establish pathway activation?
3. What if one measured gene maps to three proteins?

::: {.callout-note collapse="true" title="Suggested answers"}
1. They could not have entered the selected list under this assay.
2. No; it is evidence about representation under the supplied null.
3. Define the gene/protein analysis unit and avoid counting repeated gene evidence as independent entries.
:::

## Continue learning

[B03 Databases](../notes/b03_databases.qmd) and [B14 Pathways and networks](../notes/b14_networks.qmd).

Continue with [networks](p11_networks.qmd). The Multi-Omics pathway reuses this practical rather than creating a duplicate enrichment analysis. Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
