---
title: "Biological network construction and interpretation"
subtitle: "Practical 11: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Define a correlation edge; construct a symmetric adjacency matrix; calculate degree and inspect threshold sensitivity; separate association from regulation.

## Biological question and data

Which hypothetical proteins show correlated abundance across the same animals?

The complete log2 protein table contains a supplied shared latent factor and independent noise. A correlation network is descriptive. Animal observations are paired across proteins, and edges are not directional or causal.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Align and calculate associations

Use animals as the observations and protein features as variables.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
X <- t(d$protein)
stopifnot(identical(rownames(X),d$metadata$animal),!anyNA(X))
C <- cor(X)
round(C,2)
stopifnot(isSymmetric(C),all(abs(diag(C)-1)<1e-12))
```

Positive correlations are expected from the shared generating factor. The same pattern could arise from a shared technical factor in another dataset.

## 2. Define edges and calculate degree

Use a supplied absolute-correlation cutoff of .55. Remove self-edges and count each undirected edge once.

```{r}
cutoff <- .55
A <- abs(C)>=cutoff; diag(A) <- FALSE
edges <- which(A & upper.tri(A),arr.ind=TRUE)
degree <- rowSums(A)
data.frame(protein=colnames(X),degree)
stopifnot(isSymmetric(A),sum(degree)==2*nrow(edges))
angle <- seq(0,2*pi,length.out=ncol(X)+1)[-(ncol(X)+1)]
pos <- cbind(cos(angle),sin(angle))
plot(pos,type="n",axes=FALSE,xlab="",ylab="",asp=1,
 xlim=c(-1.35,1.35),ylim=c(-1.35,1.35),main="Correlation network")
if(nrow(edges)>0) for(i in seq_len(nrow(edges))) {
 e <- edges[i,];segments(pos[e[1],1],pos[e[1],2],pos[e[2],1],pos[e[2],2],col="grey60")
}
points(pos,pch=19,col="#1b9e77",cex=1.5)
text(pos,labels=colnames(X),pos=3,cex=.7)
```

The picture is a layout of a defined graph, not a spatial map of interacting proteins.

## 3. Examine threshold dependence

Use the same variables and animals for each threshold.

```{r}
threshold <- c(.3,.55,.75)
data.frame(threshold,edges=vapply(threshold,function(t)
 sum((abs(C)>=t)[upper.tri(C)]),integer(1)))
```

Connectivity changes with the chosen threshold; a hub designation should state that dependence.

## Interpretation and limits

Correlation can reflect diet, genotype, tissue mixture, batch or other common drivers. A network inferred from abundance is not equivalent to a curated interaction database. The threshold is supplied rather than selected to optimize a biological story; stability is assessed in the biomarker practical.

## Exercises

1. Why is the diagonal removed?
2. Could a batch shift create an apparent module?
3. Why is the sum of degrees twice the edge count?

::: {.callout-note collapse="true" title="Suggested answers"}
1. A feature’s self-correlation is not an inter-feature relationship.
2. Yes; shared technical variation can correlate many measurements.
3. Each undirected edge contributes one degree to each endpoint.
:::

## Continue learning

[B14 Pathways and networks](../notes/b14_networks.qmd).

Continue with [network and biomarker validation](p24_biomarkers.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
