---
title: "Scientific reporting and reproducible analysis records"
subtitle: "Practical 13: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Connect a question, sample definition, analysis and result; document provenance and exclusions; distinguish a numerical association from a biological conclusion.

## Biological question and data

What evidence and processing record should accompany a computational result?

The example fits a simple animal-level response model to fictional data. Diet allocation, independence and the chosen covariates are assumptions of this teaching design. The report records a conditional estimate, not a livestock recommendation.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Define and fit the analysis

Write down the target contrast before fitting. Include batch and genotype1 because the supplied model contains these patterns or sources of variation.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
meta <- d$metadata;meta$dosage <- d$genotypes[meta$animal,"marker1"]
fit <- lm(response~diet+batch+dosage,data=meta)
summary(fit)$coefficients
interval <- confint(fit)["dietsupplement",]
stopifnot(nobs(fit)==48,all(is.finite(interval)))
```

The diet coefficient adjusts for the supplied batch and genotype columns. The model still summarizes a particular hypothetical sample under its assumptions.

## 2. Create a compact evidence record

Preserve the contrast, analysis population, estimate, interval and exclusions beside the input-generation record.

```{r}
record <- data.frame(question="Diet contrast in fictional animals",unit="animal",
 n=nobs(fit),exclusions=0,estimate=unname(coef(fit)["dietsupplement"]),
 lower=interval[1],upper=interval[2],units="arbitrary response units",
 input_code_checksum=unname(tools::md5sum(file.path(helper_dir,"teaching_data.R"))))
record
```

The checksum identifies the local input-generating source. It is not a biological validity check.

## 3. Write the interpretation at the level supported

Separate result, limitation and follow-up. The generated sentence deliberately avoids a causal-mechanism claim.

```{r}
cat(sprintf("Among %d fictional animals, the adjusted diet contrast is %.2f arbitrary units (95%% model interval %.2f to %.2f).\n",
 nobs(fit),coef(fit)["dietsupplement"],interval[1],interval[2]))
```

Add the allocation, covariates, model assumptions and absence of external validation. An interval alone is not a complete report.

## Interpretation and limits

A real report needs species/population, measurement units, design, permissions, assay processing, software, missingness, diagnostics, validation and uncertainty. Tables and plots should allow readers to assess the conclusion without relying on an unexplained significant/not-significant label.

## Exercises

1. Name one claim the result cannot support.
2. What changes if ten animals are excluded after looking at their outcomes?
3. Which parts of a report remain useful when data cannot be public?

::: {.callout-note collapse="true" title="Suggested answers"}
1. A specific molecular mechanism or a dependable effect in another farm/population.
2. Selection may bias the estimate; document and justify exclusions and assess sensitivity.
3. Methods, variable definitions, provenance, permissions/access process and a non-sensitive reproducible demonstration.
:::

## Continue learning

[B17 Integrity](../notes/b17_integrity.qmd) and [B18 Synthesis](../notes/b18_synthesis.qmd).

Continue with [research proposal](p26_research_proposal.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
