---
title: "Sample identity, table joins and missing assays"
subtitle: "Practical 14: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Align assay columns to biological IDs; reject duplicate keys; distinguish absent assays from missing feature measurements; evaluate row-order errors.

## Biological question and data

What happens when protein columns arrive in a different order from phenotype metadata?

Every fixture animal has an assay identifier. The shuffled column order is deliberate; the underlying biological labels remain correct. This checks metadata alignment, not physical sample identity.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Create a plausible order mismatch

Reverse the assay order while leaving labels attached. Dimensions still match.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
meta <- d$metadata
shuffled <- d$protein[,rev(seq_len(ncol(d$protein))),drop=FALSE]
stopifnot(ncol(shuffled)==nrow(meta),!identical(colnames(shuffled),meta$animal))
head(data.frame(metadata_animal=meta$animal,assay_column=colnames(shuffled)))
```

A dimension check alone cannot identify this mistake.

## 2. Align explicitly and verify identity

Require unique IDs and complete matches before reordering.

```{r}
stopifnot(!anyDuplicated(meta$animal),!anyDuplicated(colnames(shuffled)))
idx <- match(meta$animal,colnames(shuffled))
stopifnot(!anyNA(idx))
aligned <- shuffled[,idx,drop=FALSE]
stopifnot(identical(colnames(aligned),meta$animal),
 max(abs(aligned-d$protein))<1e-12)
c(correct_cor=cor(aligned[1,],meta$response),
 wrong_order_cor=cor(shuffled[1,],meta$response))
```

The correctly aligned table reproduces the original measurements exactly. A wrong-order association can disappear or change direction.

## 3. Separate missing animals from missing features

Remove one entire assay column and compare this to NAs within otherwise present assays.

```{r}
partial <- aligned[,-1,drop=FALSE]
index <- match(meta$animal,colnames(partial))
c(animals_without_assay=sum(is.na(index)),
 missing_feature_values=sum(is.na(d$protein_missing)))
stopifnot(sum(is.na(index))==1,sum(is.na(d$protein_missing))==11)
```

An absent assay and an unobserved protein value need different handling and reporting. Neither is a measured zero.

## Interpretation and limits

ID matching preserves supplied labels but cannot detect a mislabeled specimen collected in the field. Use collection records and appropriate sample-identity checks for real data. Avoid silent complete-case deletion when missing assays relate to treatment or health.

## Exercises

1. Why does a sorted join still need an identity check?
2. What if one metadata animal has two assay columns?
3. Why is a missing protein value not log2 zero?

::: {.callout-note collapse="true" title="Suggested answers"}
1. Sorting conventions and keys can differ between tables.
2. Define technical replicates or repeated specimens explicitly; do not choose a duplicate silently.
3. Missing means unobserved; log2 zero corresponds to an original-scale value of one.
:::

## Continue learning

[B03 Databases](../notes/b03_databases.qmd), [B04 Programming](../notes/b04_programming.qmd) and [B15 Integration](../notes/b15_integration.qmd).

Continue with [multi-omics QC](p16_multiomics_qc.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
