---
title: "Multi-omics experimental design in animal production"
subtitle: "Practical 15: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Define the biological question and experimental unit; plan matched assays and balanced processing; diagnose a rank-deficient design.

## Biological question and data

How should an individually allocated diet study collect molecular profiles and later performance measurements?

The hypothetical design allocates diet to individual animals. Pen-allocated diets would require a different unit and replication plan. The practical specifies metadata and tests design identifiability; it does not establish sample size or a collection protocol for a real assay.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Build a matched sample sheet

Keep animal, diet and batch separate from the assay-specific specimen identifiers.

```{r}
plan <- data.frame(animal=paste0("a",1:12),
 diet=factor(rep(c("control","supplement"),6)),
 batch=factor(rep(c("A","A","B","B"),3)),
 tissue="liver",collection_day=14)
plan$rna_id <- paste0(plan$animal,"_RNA_d14")
plan$protein_id <- paste0(plan$animal,"_protein_d14")
stopifnot(!anyDuplicated(plan$animal),!anyDuplicated(plan$rna_id))
plan
xtabs(~diet+batch,plan)
```

Each animal has two linked assays from a specified tissue and day. The sample sheet still needs documented collection/preservation and assay-specific handling.

## 2. Check estimability

Contrast a crossed design with perfect treatment–batch confounding.

```{r}
X <- model.matrix(~diet+batch,plan)
bad <- plan;bad$batch <- bad$diet
Xbad <- model.matrix(~diet+batch,bad)
rbind(crossed=c(columns=ncol(X),rank=qr(X)$rank),
 confounded=c(columns=ncol(Xbad),rank=qr(Xbad)$rank))
stopifnot(qr(X)$rank==3,qr(Xbad)$rank==2)
```

Full rank permits separate estimation of these fixed-effect columns. It does not certify power or remove unmeasured confounding.

## 3. Plan decision timing and missingness

Specify the predictor window and outcome window before inspecting results.

```{r}
decisions <- data.frame(item=c("genotype","day14 assays","future performance"),
 available_day=c(0,14,28),role=c("baseline covariate","candidate day14 predictor","outcome"))
decisions
stopifnot(all(decisions$available_day[decisions$role=="candidate day14 predictor"]<=14))
```

The future outcome can evaluate a day14 prediction; it cannot be included as input to that prediction. Predefine how absent assays and failed QC affect the analysis population.

## Interpretation and limits

Randomization and blocking, tissue/time, sex, farm, batch, handling and exclusion rules need a justified study-specific plan. Technical aliquots are not independent animals. Use [NC3Rs design guidance](https://nc3rs.org.uk/3rs-resources/key-elements-well-designed-experiment) for the design principles; the invented sheet is not a laboratory sampling instruction.

## Exercises

1. If diet is assigned to pens, what changes?
2. Why can adding animals within the same confounded batches fail to solve the problem?
3. What must be decided before selecting an integration method?

::: {.callout-note collapse="true" title="Suggested answers"}
1. Pen-level allocation and dependence become central; sufficient independent pens are needed.
2. The treatment and batch columns remain dependent.
3. Question, samples, matching, time, measurement scales, missingness, covariates and validation target.
:::

## Continue learning

[B01 Foundations](../notes/b01_foundations.qmd), [B02 Design](../notes/b02_design.qmd) and [B18 Synthesis](../notes/b18_synthesis.qmd).

Continue with [multi-omics quality assessment](p16_multiomics_qc.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
