---
title: "Metabolomics processing, standards and feature interpretation"
subtitle: "Practical 17: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Evaluate a multiplicative-drift correction; preserve feature identity and measurement units; distinguish detected features from assigned metabolites.

## Biological question and data

How does an internal standard help distinguish instrument drift from a supplied concentration pattern?

The first calculation has an ideal shared drift factor affecting analyte and standard equally. The fixture metabolite table is already log2 arbitrary abundance; its feature labels do not identify actual compounds.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Test an ideal measurement correction

An analyte and a standard are multiplied by the same hypothetical response factor.

```{r}
concentration <- c(2,3,5,4,3,6)
drift <- c(.8,.9,1,1.1,1.2,1.3)
raw <- concentration*drift
standard <- 10*drift
corrected <- 10*raw/standard
data.frame(concentration,raw,standard,corrected)
stopifnot(max(abs(corrected-concentration))<1e-12)
```

Correction is exact only because the generating process supplies a shared multiplier and a stable standard.

## 2. Break the standard assumption

Allow analyte response to vary differently from the standard.

```{r}
analyte_drift <- drift*c(1,.95,1.1,.9,1.05,1)
biased <- 10*(concentration*analyte_drift)/standard
data.frame(concentration,estimate=biased,error=biased-concentration)
```

A standard cannot correct variation it does not track. Chemical properties and preparation can give different responses.

## 3. Explore a separate log2 feature panel

Maintain labels and scales before plotting. The annotation-confidence labels are fictional examples.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
M <- d$metabolite
identity <- data.frame(feature=rownames(M),assignment=c("hypothetical named candidate","unresolved feature",
 "unresolved feature","hypothetical named candidate","unresolved feature","unresolved feature"))
identity
boxplot(M[1,]~d$metadata$diet,xlab="Diet",ylab="Supplied log2 abundance",col="#86B7A3")
c(log2_mean_difference=mean(M[1,d$metadata$diet=="supplement"])-
 mean(M[1,d$metadata$diet=="control"]))
```

The difference describes the hypothetical panel. A mass-spectrometry feature cannot be assigned to a pathway merely because its group contrast is large.

## Interpretation and limits

Real workflows need blanks, technical QC, drift and batch assessment, detection/identification confidence and justified normalization. The [EMBL-EBI metabolite-identification material](https://www.ebi.ac.uk/training/materials/metabolomics-analysis-materials/methods-and-resources/metabolite-identification/) explains why chemical assignment is separate from signal detection.

## Exercises

1. Why did the imperfect correction fail?
2. Would another log transform be appropriate for M?
3. Can an unresolved feature support a named pathway claim?

::: {.callout-note collapse="true" title="Suggested answers"}
1. The standard did not share the same response factor as the analyte.
2. No; M already represents log2 abundance.
3. Not without justified chemical identification or carefully qualified evidence.
:::

## Continue learning

[B09 Metabolomics](../notes/b09_metabolomics.qmd) and [B12 Statistics](../notes/b12_statistics.qmd).

Continue with [exploration](p19_exploration.qmd) or [integration](p21_integration.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
