---
title: "Multi-omics integration with matched samples and balanced blocks"
subtitle: "Practical 21: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Align layers; compare unit scaling and block weighting; inspect a joint representation; distinguish shared latent variation from independent confirmation.

## Biological question and data

What changes when protein and metabolite layers are integrated rather than inspected separately?

The 48 animals have complete matched log2 measurements in these two layers. A shared latent factor was deliberately used to generate them. PCA is descriptive; its full-sample transformation is not used as a prediction evaluation.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Align layers before integration

Keep biological IDs visible and check them against metadata.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
P <- t(d$protein);M <- t(d$metabolite)
stopifnot(identical(rownames(P),d$metadata$animal),identical(rownames(P),rownames(M)))
c(protein_features=ncol(P),metabolite_features=ncol(M),animals=nrow(P))
```

An assay match requires the right animal and sampling context, not just equal matrix dimensions.

## 2. Scale features and weight blocks

Standardize each feature. Then divide each block by the square root of its feature count so its total marginal variance is one.

```{r}
ZP <- scale(P);ZM <- scale(M)
balanced <- cbind(ZP/sqrt(ncol(P)),ZM/sqrt(ncol(M)))
c(protein_total_variance=sum(apply(ZP/sqrt(ncol(P)),2,var)),
 metabolite_total_variance=sum(apply(ZM/sqrt(ncol(M)),2,var)))
stopifnot(abs(sum(apply(ZP/sqrt(ncol(P)),2,var))-1)<1e-12,
 abs(sum(apply(ZM/sqrt(ncol(M)),2,var))-1)<1e-12)
fit <- prcomp(balanced,center=FALSE,scale.=FALSE)
head(fit$sdev^2/sum(fit$sdev^2))
```

Equal total marginal variance is a specified weighting choice. It does not assert equal information, independence or biological importance.

## 3. Inspect the representation and shared association

Label scores by the biological design and calculate one paired cross-layer association.

```{r}
plot(fit$x[,1],fit$x[,2],col=as.integer(d$metadata$diet),pch=19,
 xlab="Joint PC1",ylab="Joint PC2")
legend("topleft",levels(d$metadata$diet),col=1:2,pch=19,bty="n")
c(cross_layer_correlation=cor(P[,1],M[,1]))
```

The panels can agree because they share the same generating driver. In actual data, biological and technical common causes both need examination.

## Interpretation and limits

Concatenation is one integration strategy, not a universal method. Unmatched samples, incompatible timing, missing layers and leakage need explicit handling. Do not interpret joint PCs as mechanisms or use full-sample scaling for an untouched predictive test.

## Exercises

1. What happens without the block weights?
2. Does matching animal IDs establish matching tissue/time?
3. Are correlated layers independent replications?

::: {.callout-note collapse="true" title="Suggested answers"}
1. The eight standardized protein features contribute more marginal variance than the six metabolite features.
2. No; specimen metadata must confirm context.
3. Not generally; shared samples, drivers and processing can create dependence.
:::

## Continue learning

[B15 Integration](../notes/b15_integration.qmd), [B02 Design](../notes/b02_design.qmd) and [B12 Statistics](../notes/b12_statistics.qmd).

Continue with [integrated predictive analysis](p23_integrated_analysis.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
