---
title: "Multi-omics associations with an animal-production trait"
subtitle: "Practical 22: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Estimate conditional feature associations; compare marginal and adjusted patterns; report effect scale and multiple testing; distinguish prediction from mechanism.

## Biological question and data

Which hypothetical molecular features are associated with the supplied animal response after accounting for diet and batch?

Animals are independent in this fixture. Response units are arbitrary; features are already log2 abundance. The shared latent factor generates association without specifying measured features as causes. Linear-model inference remains conditional on the selected covariates and residual assumptions.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Prepare matched measurements and the design

Keep animal rows aligned and examine the crossed design.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
X <- cbind(t(d$protein),t(d$metabolite));meta <- d$metadata
stopifnot(identical(rownames(X),meta$animal))
xtabs(~diet+batch,meta)
summary(meta$response)
```

The target is a feature association conditional on measured covariates, not a treatment effect or breeding-value estimate.

## 2. Fit a prespecified conditional model per feature

For each feature use response ~ feature + diet + batch. Report the association per one log2 unit of the feature.

```{r}
analyse <- function(feature) {
 z <- meta;z$feature <- feature
 fit <- lm(response~feature+diet+batch,z)
 tab <- coef(summary(fit))
 c(effect=tab["feature","Estimate"],SE=tab["feature","Std. Error"],
 p=tab["feature","Pr(>|t|)"])
}
result <- as.data.frame(t(apply(X,2,analyse)))
result$BH <- p.adjust(result$p,"BH")
result[order(result$p),]
stopifnot(all(is.finite(as.matrix(result))),all(result$BH>=result$p-1e-12))
```

The adjusted p-values concern the supplied set of tests under the model and applicable dependence conditions. They do not identify causal biomarkers.

## 3. Compare an unadjusted and conditional relationship

Use the same feature and animals for both fits.

```{r}
z <- meta;z$feature <- X[,1]
a <- lm(response~feature,z)
b <- lm(response~feature+diet+batch,z)
c(marginal=coef(a)["feature"],conditional=coef(b)["feature"])
plot(z$feature,z$response,pch=19,col=as.integer(z$diet),
 xlab="Feature log2 abundance",ylab="Response (arbitrary units)")
```

Adjustment changes the estimand; it is not an automatic guarantee that every nuisance variable has been removed.

## Interpretation and limits

Marker relatedness, repeated animals, nonlinear associations, measurement error and selection require suitable analysis in real data. A feature selected on all outcomes cannot then be evaluated on those same animals as an independently validated predictor.

## Exercises

1. What units does the fitted feature coefficient have?
2. Why can a feature predict without being a cause?
3. Where should feature selection occur in a prediction workflow?

::: {.callout-note collapse="true" title="Suggested answers"}
1. Arbitrary response units per one log2 unit of the supplied abundance.
2. It may track a shared biological or technical driver.
3. Within training-only validation; then evaluate the complete selected pipeline on held-out animals.
:::

## Continue learning

[B12 Statistics](../notes/b12_statistics.qmd), [B13 Association](../notes/b13_association.qmd) and [B18 Synthesis](../notes/b18_synthesis.qmd).

Continue with [prediction](p20_prediction.qmd) or [biomarker validation](p24_biomarkers.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
