---
title: "Multi-omics research proposal and scientific reporting"
subtitle: "Practical 26: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Design a coherent study from question to reporting; specify primary outcomes, unit, sampling and analysis; justify each assay and validation step.

## Biological question and data

How would you propose a study of dietary supplementation and later feed efficiency while distinguishing response, prediction and mechanism?

This is a design practical, not an animal-use protocol or a power calculation. The illustrative budget uses arbitrary cost units. Replace these with justified estimates and appropriate approvals in a real project.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Specify the estimand and sampling plan

Write a primary question and separate secondary objectives. Avoid promising all three with one unqualified association analysis.

```{r}
proposal <- data.frame(field=c("primary question","experimental unit","sampling","primary outcome","secondary analysis"),
 decision=c("diet contrast in later efficiency","individually allocated animal",
 "matched day14 RNA/protein; documented handling",
 "predefined days15–28 efficiency with calibrated units",
 "exploratory profile association; independent prediction evaluation"))
proposal
```

The primary contrast, sampling time and outcome window are explicit. Tissue, farm/pen structure, allocation, sample size and feasibility still need specification.

## 2. Make resource trade-offs visible

Calculate a transparent illustrative assay budget without treating it as a sample-size justification.

```{r}
cost <- c(RNA=3,protein=4,metabolite=2)
animals <- 24
options <- data.frame(design=c("RNA only","RNA plus protein","three layers"),
 assay_cost=animals*c(cost["RNA"],sum(cost[c("RNA","protein")]),sum(cost)))
options
stopifnot(options$assay_cost[3]==216)
```

A cheaper design can support more biological replication or better validation. Cost arithmetic alone cannot identify the statistically or biologically best design.

## 3. Define the analysis and reporting deliverable

Require each step to connect to the question and assumptions.

```{r}
deliverable <- data.frame(section=c("design","data/QC","analysis","validation","interpretation","governance"),
 requirement=c("unit, allocation, covariates and sampling",
 "IDs, handling, assay scale, missingness and exclusions",
 "prespecified contrasts and justified integration",
 "independent groups and training-only choices",
 "effect/uncertainty, alternatives and follow-up",
 "access, provenance, sharing and responsible AI"))
deliverable
```

Assess a proposal for coherence: the methods should answer the stated question with measurements available at the right time.

## Interpretation and limits

Formal sample size, actual assay protocols, permits, data agreements, model qualification and collaboration-specific assessment rubrics are outside this invented proposal. The original course assessment weights remain in the curriculum map.

## Exercises

1. Why should a primary outcome be specified?
2. What is sacrificed when every possible omics layer is added with a fixed budget?
3. How would a computational-report extension differ from this proposal?

::: {.callout-note collapse="true" title="Suggested answers"}
1. It anchors design, sample-size reasoning and interpretation instead of selecting the most favourable result.
2. Potentially replication, assay quality, follow-up or independent validation.
3. It would implement and evaluate the prespecified pipeline, report exclusions/results/uncertainty and demonstrate reproducibility rather than only promise them.
:::

## Continue learning

[B02 Design](../notes/b02_design.qmd), [B17 Integrity](../notes/b17_integrity.qmd) and [B18 Synthesis](../notes/b18_synthesis.qmd).

Review the [reporting practical](p13_reporting.qmd) and choose a pathway in the [course map](../course_map.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
