---
title: "Microbiome composition and host–microbiome interpretation"
subtitle: "Practical 27: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Distinguish absolute abundance, read counts and relative composition; calculate a positive-component log ratio; interpret host association without implying mechanism.

## Biological question and data

Can a taxon’s relative decrease occur without any absolute decrease, and how should a microbial association with performance be interpreted?

The first table supplies invented absolute quantities. The shared fixture then supplies fictional read counts; these do not identify absolute microbial load or activity. A .5 pseudocount is used only for a sensitivity demonstration, not as a universally valid zero solution.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Compare absolute and relative changes

Increase one component while keeping the others fixed.

```{r}
absolute <- rbind(before=c(A=10,B=10,C=10),after=c(A=20,B=10,C=10))
relative <- absolute/rowSums(absolute)
relative
stopifnot(absolute[1,"B"]==absolute[2,"B"],relative[2,"B"]<relative[1,"B"])
```

B decreases in proportion because A increased. The fraction alone cannot diagnose absolute depletion.

## 2. Inspect a hypothetical read panel

Normalize within each sample and verify that fractions sum to one.

```{r}
helper_dir <- if (dir.exists("R")) "R" else "../R"
stopifnot(file.exists(file.path(helper_dir,"teaching_data.R")))
source(file.path(helper_dir,"teaching_data.R"))
d <- teaching_data()
counts <- d$microbial
fraction <- sweep(counts,2,colSums(counts),"/")
stopifnot(max(abs(colSums(fraction)-1))<1e-12,
 identical(colnames(fraction),d$metadata$animal))
barplot(fraction[,1:8],col=rainbow(nrow(fraction)),ylab="Relative read fraction",
 xlab="Fictional animals",legend.text=rownames(fraction))
```

Relative read representation combines biology and measurement. Extraction and assay properties affect what is observed.

## 3. Examine a log ratio and zero sensitivity

Use taxon01/taxon02 with explicit pseudocount values. A ratio removes a common multiplicative sample factor for positive components.

```{r}
ratio <- log((counts[1,]+.5)/(counts[2,]+.5))
c(response_correlation=cor(ratio,d$metadata$response))
zero_case <- c(A=0,B=10)
pseudocount <- c(.1,.5,1)
data.frame(pseudocount,log_ratio=log((zero_case["A"]+pseudocount)/(zero_case["B"]+pseudocount)))
```

The zero-case ratio depends materially on the pseudocount. A host correlation in this panel is a descriptive association, not a microbial intervention result.

## Interpretation and limits

Distinguish amplicon/shotgun methods, site, time, diet, contaminants, compositionality and load. The fixture does not model a real rumen community. Cross-layer network or methane claims require identified measurements and suitable designs.

## Exercises

1. Why cannot fractions recover total microbial load?
2. What assumption is hidden in a pseudocount?
3. What could confound a taxon–production association?

::: {.callout-note collapse="true" title="Suggested answers"}
1. Multiplying all absolute quantities by the same factor preserves fractions.
2. It supplies an artificial positive value for an unobserved/zero component, whose effect depends on scale and detection process.
3. Diet, management, host genotype, sampling site/time and technical processing.
:::

## Continue learning

[B10 Microbiomics](../notes/b10_microbiomics.qmd), [B14 Networks](../notes/b14_networks.qmd) and [B15 Integration](../notes/b15_integration.qmd).

Continue with [integrated interpretation](p25_interpretation.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
