---
title: "Epigenomic coverage and cell-composition effects"
subtitle: "Practical 30: Genomics, systems biology and multi-omics"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
  warning: false
---

## Learning objectives

Calculate methylation proportions with explicit read denominators; distinguish measurement precision from biological replication; inspect a composition-driven group difference.

## Biological question and data

Can a bulk-tissue methylation difference arise with no within-cell-type regulatory change?

Two fictional cell types have constant methylation probabilities .2 and .8. Group-specific cell mixture determines the pooled probability. Reads are binomial conditional on that probability; this simplified model does not represent all assay or biological variation.

No biological download is required. Numerical inputs are hypothetical, not observed animal data. The runnable examples use base R. Count and abundance units are specified where used.

## 1. Construct the mixture model

Change cell proportions without changing either cell type’s methylation.

```{r}
cell_methylation <- c(typeA=.2,typeB=.8)
mixture <- rbind(control=c(.8,.2),supplement=c(.3,.7))
probability <- as.vector(mixture%*%cell_methylation)
names(probability) <- rownames(mixture)
probability
stopifnot(abs(probability[1]-.32)<1e-12,abs(probability[2]-.62)<1e-12)
```

The expected bulk difference is .30, caused entirely by composition in this hypothetical model.

## 2. Sample reads from independent animals

Read depth changes conditional measurement precision but not the number of animals.

```{r}
set.seed(3001)
group <- rep(c("control","supplement"),each=6)
depth <- rep(c(10,20,40,80,10,20),2)
truth <- probability[group]
methylated <- rbinom(12,depth,truth)
unmethylated <- depth-methylated
observed <- methylated/depth
out <- data.frame(animal=paste0("a",1:12),group,depth,methylated,unmethylated,observed,
 conditional_SE=sqrt(truth*(1-truth)/depth))
out
stopifnot(all(methylated+unmethylated==depth),all(observed>=0),all(observed<=1))
```

Twelve animals remain twelve animals even if their read counts are large. The supplied conditional SE assumes the true mixture probability is known.

## 3. Compare animal and read-weighted summaries

A mean of animal proportions and a pooled read fraction use different weighting.

```{r}
animal_mean <- tapply(observed,group,mean)
read_weighted <- tapply(methylated,group,sum)/tapply(depth,group,sum)
data.frame(group=names(animal_mean),animal_mean=as.vector(animal_mean),
 read_weighted=as.vector(read_weighted))
boxplot(observed~group,xlab="Group",ylab="Observed methylated-read fraction",col="#86B7A3")
```

Pooling reads gives deeper samples more weight. It is not automatically the correct estimator of an average-animal contrast.

## Interpretation and limits

Coverage, conversion efficiency, mapping, tissue/cell composition and assay specificity must be evaluated in actual studies. A group contrast in bulk methylation does not establish a within-cell regulatory mechanism or phenotypic mediation.

## Exercises

1. Did the cell-type probabilities change?
2. What does doubling depth improve under this model?
3. Why is a pooled-read contrast not a new treatment replication?

::: {.callout-note collapse="true" title="Suggested answers"}
1. No; only their proportions changed.
2. Conditional sampling precision for each animal’s fraction, not biological replication or control of confounding.
3. Reads are nested measurements of existing animals; independent allocations determine treatment replication.
:::

## Continue learning

[B07 Epigenomics](../notes/b07_epigenomics.qmd), [B02 Design](../notes/b02_design.qmd) and [B15 Integration](../notes/b15_integration.qmd).

Continue with [advanced measurement units](p31_advanced.qmd). Return to the [practical index](../practicals.qmd) or the [unified map](../course_map.qmd).
